Acceleration of Bilateral filtering algorithm for manycore and multicore architectures

### 2012-03-10 trunk/LateX/paper8.tex
more minor changes

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images. As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e. one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration. A formal expression is given below:

@@

In addition to restating the Bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the Bilateral filter. Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated. Consequently, the Bilateral filter has two data dependencies. As explained later in this paper, the pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.

@@

CUDA's intra-warp implicit synchronization capability is the cornerstone of our CUDA implementation of our pair-symmetric algorithm. Without implicit synchronization, our pair-symmetric algorithm would suffer significantly over CUDA because it would require expensive synchronization calls per Center-Neighbor Filter computation.

@@

To take full advantage of CUDA's implicit intra-warp synchronization feature, divergent branching between threads in the same warp must be avoided. This is so because all the threads in the same warp share only one program counter. A divergent branch would cause some or most of the threads to be idle as the program counter runs through the branch and drives only those threads that meet the branch condition to work. An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.

Divergent branching commonly occur in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined. In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*jMax + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.

>>>>

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image. Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration. A formal expression is given below:

@@

In addition to restating the Bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the Bilateral filter. Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated. Consequently, the Bilateral filter has two data dependencies. As explained later in this paper, our pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.

@@

CUDA's intra-warp implicit synchronization capability is the cornerstone of our CUDA implementation of our pair-symmetric algorithm. Without implicit synchronization, the performance of our CUDA-implemented pair-symmetric algorithm would be significantly reduced because it would require expensive synchronization calls per Center-Neighbor Filter computation.

@@

To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided. This is so because all the threads in the same warp share only one program counter. A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that met the branch condition to work. An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.

Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined. In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*width + j] representation of a 2D image, for if threads in a warp were to read from or write to array elements (representing pixels) in a vertical order, there would be a separate transaction for each thread in the warp.

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work should be arranged such that warps can work as independently from each other as possible in order to minimize explicit synchronization calls, as illustrated in Figure~\ref{fig:pspro}. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work of each warp on its current row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call would result in warps trespassing into the territory of other warps, thereby causing race conditions as threads from different warps would be consequently prone to writing to the same memory location.
